---
title: "An R package for estimating a joint model of correlated ordinal and continuous variables"
author: "Laura Vana and Rainer Hirk"
output: pdf_document
header-includes:
  - \usepackage{bm}
  - \usepackage{pdfpages}
  - \usepackage{hyperref}
  - \usepackage{xcolor}
  - \usepackage{booktabs}
  - \usepackage{multicol}
  - \newcommand{\proglang}[1]{\textbf{\texttt{#1}}}
  - \newcommand{\pkg}[1]{\textbf{#1}}
  - \newcommand{\code}[1]{\texttt{#1}}
  - \hypersetup{colorlinks=true, citecolor=blue, linkcolor=blue, urlcolor=magenta}
bibliography: refs.bib
---

# Introduction

The analysis of mixed type responses is a relevant topic in various fields of
research where one often jointly collects various continuous, binary
or ordinal outcomes possibly  with some observations missing. 

The joint modeling of outcomes is sometimes preferred over a separate 
analysis of the different responses given that the association between 
different outcomes can be captured is capture by the joint model
specification. The advantage of this approach is that answers to various 
research questions can be obtained in one go, thus eliminating the need for
multi-step procedures to combine results  from various separate analyses.


# Model      {#sect:model}

We propose a joint model of continuous and ordinal variables. Assume we 
have $q_n$ continuous responses and $q_o$ ordinal responses (in total $q=q_n+q_o$).

For the continuous responses $\boldsymbol y^n_{1},\ldots, \boldsymbol y^n_{q_n}$ 
we assume the following regression model: 
$$
y^n_{ij} = \beta_{0j} + \bm\beta_j^\top \bm x_i + \sigma_j \epsilon^n_{ij}
$$
for all observations $i=1,\ldots,n$ and $j=1,\ldots,q_n$ where 
$\bm x_i$ is a $(p\times 1)$ vector of covariates, $\bm\beta_j$ is a $(p\times 1)$ vector of response-specific regression coefficients, $\sigma_j$ is a response-specific scale parameter and $\epsilon^n_{ij}$ is a mean-zero error term.

For the ordinal variables  $\boldsymbol y^o_{1},\ldots, \boldsymbol y^o_{q_o}$ a regression model is assumed on the latent variable $\tilde{\bm y}$ underlying the ordinal observations:
$$
\tilde y^o_{ij} =\bm \beta_j^\top \bm x_i + \epsilon^o_{ij},
$$
where the observed responses are obtained by slotting the continuous latent variables using appropriate threshold parameters.
$$
y^o_{ij} = r \Rightarrow \theta_{j, r-1}\leq \tilde y^o_{ij}\leq \theta_{j, r}, \quad r\in \{1, \ldots, K_j\} \quad  -\infty\equiv\theta_{j, 0} <\theta_{j, 1}<\ldots<\theta_{j, K}\equiv \infty,
$$
where $K_j$ denotes the number of categories for ordinal response $j=1, \ldots, q_o$.

The dependence is captured by assuming that the errors of the continuous and the latent variables
of the ordinals come from a multivariate normal distribution:
$$
(\boldsymbol\epsilon^o_{i}, \boldsymbol\epsilon^n_{ij})^\top \sim N(\bm 0, R)
$$
with mean zero and correlation matrix $R$. Note that the matrix is restricted to have diagonal elements equal to one due to identifiability restrictions for the ordinal variables.

# Pairwise likelihood {#sect:pl}

Assume we collect all $q$ responses in a matrix $Y$ of dimension
$(n\times q)$ and that there are no missing values in the response
matrix.
The pairwise log-likelihood is given by the sum of the bivariate likelihoods over all pairs of responses $k=1,\ldots,q$ and $l > k$:
$$
p\ell(\Theta; Y, X) =\sum_{i=1}^n \sum_{k<l} \ell(\Theta; y_{ik}, y_{il}).
$$
The bivariate log-likelihoods can be split into three cases, depending on the type of the responses $k$ and $l$: 

* Case 1: both responses are ordinal

* Case 2: both responses are normal

* Case 3: one response is ordinal, one normal


For *Case 1*, the bivariate log-likelihood is given by the bivariate probability $\Pr(y_{ik}=r_{ik}, y_{il} = r_{il})$:
$$
\ell(\Theta; y_{ik}, y_{il})=\Pr(y_{ik}=r_{ik}, y_{il} = r_{il}) = 
\int_{\theta_{k, r_{ik}-1}}^{\theta_{k, r_{ik}}} 
\phi_2(\tilde y_{ik} -\bm \beta_k^\top \bm x_i, \tilde y_{il} - \bm\beta_k^\top \bm x_i; \rho_{kl}) d\tilde y_{ik}d\tilde y_{il},
$$
where $\rho_{kl}$ is the correlation coefficient between responses $k$ and $l$ and
$\phi_2$ is
the bivariate standard normal probability density function. 
By denoting
$U_{i,k} = \theta_{k, r_{ik}} -\bm \beta_k^\top \bm x_i$ ,
$L_{i,k} = \theta_{k, r_{ik}-1} - \bm \beta_k^\top \bm x_i$,
$U_{i,l} = \theta_{k, r_{il}} - \bm \beta_l^\top \bm x_i$,
$L_{i,l} = \theta_{k, r_{ik}-1} - \bm \beta_l^\top \bm x_i$, 
the bivariate probability is equal to:
$$
\Pr(y_{ik}=r_{ik}, y_{il} = r_{il})=\Phi_2\left(U_{i,k}, U_{i,l};  \rho_{kl} \right)-\Phi_2\left(L_{i,k}, U_{i,l};  \rho_{kl} \right)-\Phi_2\left(U_{i,k}, L_{i,l};  \rho_{kl} \right)
+\Phi_2\left(L_{i,k}, L_{i,l};  \rho_{kl} \right)
$$
where $\Phi_2$ is the bivariate standard normal cumulative distribution function.

For *Case 2*, where both responses are continuous, the bivariate log-likelihood is given by the bivariate standard normal density:
$$
\ell(\Theta; y_{ik}, y_{il}) = f(y_{ik}, y_{il}|\Theta)= \phi_2\left(\frac{y_{ik} - \beta_{0k} -\bm\beta_k^\top \bm x_i}{\sigma_k}, \frac{y_{il} - \beta_{0l} -\bm\beta_l^\top \bm x_i}{\sigma_l}; \rho_{kl}) \right)
$$

Finally, for *Case 3*, where one response is ordinal and one response is continuous, the bivariate log-likelihood is given 
by the product of a conditional probability and the  density
of the continuous response. Assuming that the $k$-th response is ordinal and
the $l$-th is continuous, we have:
$$
\ell(\Theta; y_{ik}, y_{il}) = \Pr(y_{ik}=r_{ik} | y^n_{il}) f(y_{il})
$$
where $f(y_{il})$ is the marginal standard normal pdf $\phi\left(\frac{y^n_{il} - \beta_{0l} -\bm\beta_l^\top \bm x_i}{\sigma_l} \right)$. 
The conditional probability is given by :
$$
\Pr(y_{ik}=r_{ik} | y_{il}) = \Pr(\theta_{k, r_{ik}-1}\leq \tilde y_{ik}\leq \theta_{k, r_{ik}} | y_{il}).
$$
Using the fact that  $\tilde y_{ik}$ and $y_{il}$ are bivariate normals, we can easily derive the conditional distribution
$$
\tilde y_{ik} | y_{il} \sim N\left(\underbrace{\bm\beta_k^\top \bm x_i + \frac{\rho_{kl}}{\sigma_l}(y_{il} - \beta_{0l} -\bm\beta_l^\top \bm x_i)}_{\mu_c}, \underbrace{(1 - \rho_{kl}^2)}_{\sigma_c^2}\right)
$$
and the corresponding conditional probability:
$$
\Pr(y_{ik}=r_{ik} | y_{il})=
\int_{\theta_{k, r_{ik}-1}}^{\theta_{k, r_{ik}}} \phi\left(\frac{\tilde y_{ik} - \mu_c}{\sigma_c}\right) d\tilde y_{ik} = 
\Phi\left(\frac{\theta_{k, r_{ik}} - \mu_c}{\sigma_c}\right)-\Phi\left(\frac{\theta_{k, r_{ik}-1} - \mu_c}{\sigma_c}\right).
$$

The standard errors of the parameters we computed using the Godambe information 
matrix  and model comparison can be performed using modified version of Akaike and 
Bayesian information criteria [for more details see @pub:mvord:Hirk+Hornik+Vana:2020]

###### Missing values
Note that the framework is easily extended to allow for the presence of 
missing values in the responses. For each observation $i$, in the calculation
of the pairwise log-likelihood, we consider all
pairs of *observed responses*. In case for some observations only one response is 
available, we consider the likelihood of the univariate observed response.
For the continuous responses this is the standard normal pdf $\phi\left(\frac{y^n_{ik} - \beta_{0k} -\bm\beta_k^\top \bm x_i}{\sigma_l} \right)$ while for the ordinal 
responses we have the univariate probability 
$\Pr(y_{ik}=r_{ik})=\Phi(U_{i,k}) - \Phi(L_{i,k})$

# Software implementation {#sect:soft}

The model is implemented in package **mvordnorm**.
The main function for fitting the models is `mvordnorm()`.
```{r eval=FALSE}
mvordnorm(formula, data, response_types = NULL, na.action,
  contrasts = NULL, control = mvordnorm.control(), ...)
```

The main arguments are:

* `formula`: a formula object of class `"Formula"`, as implemented in 
   the **Formula** package [@jssFormula], which makes it easy to specify
   multiple responses, especially if these are not numeric or of mixed type.

* `data`: a data frame which contains the different responses in separate columns.

* `response_types`:  a (named) vector of characters with length equal to the number of responses. Each element of the vector is either `"gaussian"` or `"ordinal"`.

* `na.action`: a function which indicates what should happen when the data contain NAs.

* `contrasts`: an optional list. See the `contrasts.arg` of `model.matrix.default`.

* `control`: list of parameters for controlling the fitting process such as the 
general purpose solver to be used or whether standard errors should be calculated.

For illustration purposes in section we use a worked example based on a simulated data set consisting of
1000 subjects for which two multiple ordinal responses
(`y1` and `y2`), two continuous responses
(`z1` and `z2`) and three covariates (`x1`, `x2` and `x3`) are available. The ordinal responses each
have three categories labeled with 1, 2 and 3.

```{r echo=TRUE}
library("mvordnorm")
data("data_toy", package = "mvordnorm")
dim(data_toy)
head(data_toy)
```

In order to estimate the model using the CG solver in **optimx** [@jssoptimx], 
we have the following function call:
```{r}
fit <- mvordnorm("y1 + y2 + z1 + z2 ~ 0 + X1 + X2 + X3", data = data_toy,
          response_types = c("ordinal", "ordinal","gaussian", "gaussian"),
          control = mvordnorm.control(se = TRUE, solver = "CG"))
```
Note that the formula specifies no intercept is to be estimated, this is however
due to the intercept not being identifiable in ordinal regression. For the normal 
variables, an intercept will be computed by default. This peculiarity of the 
implementation should be improved in future versions of the package, to allow the
user to specify outcome specific intercepts.

The `summary` method produces an output which is similar to the one of most
regression models, and contains information on the thresholds for ordinal responses and intercepts for the continuous response, on the outcome specific regression
coefficients, on the scale parameters for the continuous variables and the correlation
matrix $R$.
```{r}
summary(fit)
```
The `summary` method produces an output which is similar to the one of most
regression models. 

If a model should be fit to data containing missing values, the `na.action` argument 
should be set to `na.pass`. We introduce some NAs in `data_toy`:
```{r}
data_toy$y1[sample(1:nrow(data_toy), 20)] <- NA
data_toy$y2[sample(1:nrow(data_toy), 20)] <- NA
```
The function call is:
```{r}
fit_with_NAs <- mvordnorm("y1 + y2 + z1 + z2 ~ 0 + X1 + X2 + X3", data = data_toy,
                          na.action = na.pass,
                          response_types = c("ordinal", "ordinal","gaussian", "gaussian"),
                          control = mvordnorm.control(se = TRUE, solver = "CG"))
summary(fit_with_NAs)
```

#### Acknowledgments
This research was supported by funds of the Oesterreichischen Nationalbank
(Austrian Central Bank, Anniversary Fund, project number: 18482 ``Multivariate ordinal regression models for enhanced credit risk  modeling'').

# References
<div id="refs"></div>

